---
title: "Benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(sbmR)
library(glue)
library(tidyverse)
```

```{r, eval = FALSE}
library(sbmR)
library(glue)
library(tidyverse)
```

## Setup

Before we can run benchmarks we need to simulate some data and get a chain starting point using the agglomerative merging algorithm...

```{r}
set.seed(42)

group_info <- dplyr::tribble(
  ~group, ~n_nodes,
     "a",       20,
     "b",       22,
     "c",       25
)

connection_propensities <- dplyr::tribble(
 ~group_1, ~group_2, ~propensity,
      "a",      "a",         0.8,
      "a",      "b",         0.2,
      "a",      "c",         0.3,
      "b",      "b",         0.9,
      "b",      "c",         0.15,
      "c",      "c",         0.4,
)

# Simulate network
network <- sim_sbm_network(group_info, connection_propensities, edge_dist = purrr::rbernoulli)

visualize_network(network, width = '100%')
```

```{r}
my_sbm <- create_sbm(network)

my_sbm$EPS


copy_sbm <- function(old_sbm){
  new_sbm <- create_sbm(
    nodes = my_sbm$get_state(),
    edges = my_sbm$get_edges(),
    eps = old_sbm$EPS
  )
}



```

Initialize SBM object from simulated network and run collapsing algorithm to find optimal initial partitioning..

```{r}
collapse_results <- create_sbm(network) %>% 
  collapse_groups(exhaustive = TRUE, 
                  num_group_proposals = 8, 
                  num_mcmc_sweeps = 15)
```


## Timing

Now that we have our network setup we can run our benchmarking tests.

```{r}
run_mcmc_sweeps <- function(n_sweeps, just_init = FALSE){
  # Re-initialize network with new epsilon (old epsilon was 0.1) and reload the collapsed state
  my_sbm <- create_sbm(network, eps = 0.01) %>% 
    choose_best_collapse_state(collapse_results, heuristic = 'lowest')
  
  if(just_init) return(NULL)
  # Run sweeps
  purrr::map(1:n_sweeps, ~mcmc_sweep(my_sbm))
}

benchmark_sweep <- function(n_sweeps, just_init){
   bench::mark(
    run_mcmc_sweeps(n_sweeps, just_init = just_init), 
    filter_gc = FALSE, 
    min_iterations = 15)
}
n_sweeps <- 500
full_runs <- benchmark_sweep(n_sweeps = n_sweeps, just_init = FALSE)
init_runs <- benchmark_sweep(n_sweeps = n_sweeps, just_init = TRUE)


# Subtract our initialization timing from the full timing to isolate the portion
# of time from sweeps
median_time <- full_runs$median - init_runs$median
```

Median time for `r n_sweeps` sweeps is __`r median_time`__, or approximately __`r median_time/n_sweeps` per sweep.__



## Agglomerative Merging Algorithms


```{r}
library(tidyverse)
# library(furrr)
library(sbmR)


n_groups <- 4    # Four total groups
group_size <- 40 # W/ 50 nodes in each
low_num <- 2
high_num <- 10
sigma <- 4
num_group_proposals <- 5
num_mcmc_sweeps <- 5

set.seed(42)
network <- sim_basic_block_network(
  n_groups = n_groups,     
  n_nodes_per_group = group_size,  
  return_connection_propensities = TRUE
)

my_sbm <- create_sbm(network)
```

```{r}
library(tictoc)
tic()
collapse_run(my_sbm, end_group_num = 10, sigma = 5)
toc()

tic()
collapse_run(my_sbm, end_group_num = 10, sigma = 5, parallel = TRUE)
toc()
```



```{r}
library(furrr)
future::plan(future::multiprocess)

collapse_to_desired_num <- function(desired_num){
  create_sbm(network) %>% 
  collapse_groups(desired_num_groups = desired_num, 
                  sigma = sigma,
                  num_group_proposals = num_group_proposals,
                  num_mcmc_sweeps = num_mcmc_sweeps) %>% 
    select(-state) %>% 
    mutate(desired_num = desired_num)
}

r_sequential <- function(){
  purrr::map_dfr(
    low_num:high_num,
    collapse_to_desired_num
  )
}

r_parallel <- function(){
  future_map_dfr(
    low_num:high_num,
    collapse_to_desired_num
  )
}

pkg_sequential <- function(){
  collapse_run(my_sbm, 
               start_group_num = low_num,
               end_group_num = high_num, 
               sigma = sigma, 
               num_group_proposals = num_group_proposals,
               num_mcmc_sweeps = num_mcmc_sweeps,
               parallel = FALSE)
}

pkg_parallel <- function(){
  collapse_run(my_sbm, 
               start_group_num = low_num,
               end_group_num = high_num, 
               sigma = sigma, 
               num_group_proposals = num_group_proposals,
               num_mcmc_sweeps = num_mcmc_sweeps,
               parallel = TRUE)
}


timing_results <- bench::mark(
  r_sequential = r_sequential(),
  r_parallel = r_parallel(),
  pkg_sequential = pkg_sequential(),
  pkg_parallel = pkg_parallel(),
  check = FALSE,
  min_iterations = 5
)


plot(timing_results)
```


