---
title: "agglomerative_merging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{agglomerative_merging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One technique of finding the optimal number of clusters/ best partitioning is using the efficient agglomerative merging algorithm. This is typically used to find the initial state for MCMC chains but can stand on its own as a clustering method. 

Agglomerative merging can be acomplished with the SBM class using the method `collapse_groups()`. 


First we will setup a model using a simulated bipartite network dataset. The true number of groups in this dataset is __6__.

```{r setup}
library(bisbmsim)
library(sbmR)
library(tidyverse)
```


```{r}
N_a <- 50  # Number of nodes of the a type
N_b <- 50  # Number of nodes of the b type
K_a <- 5    # How many blocks of a type nodes are there
K_b <- 5    # How many blocks of the b type nodes are there
b_a <- assign_group_membership(N = N_a, K = K_a) # Block membership for each a node
b_b <- assign_group_membership(N = N_b, K = K_b) # Block membership for each b node


edges <- draw_from_model(
  b_a, b_b,
  Lambda = generate_random_lambda(K_a = K_a, K_b = K_b)
) %>%
  filter(num_edges > 0) %>%
  select(a, b) %>%
  transmute(
    from = paste0('a',a),
    to = paste0('b', b)
  )

nodes <- dplyr::bind_rows(
  tibble(
    id = unique(edges$from),
    type = "a"
  ),
  tibble(
    id = unique(edges$to),
    type = "b"
  )
)


# Setup SBM model
my_sbm <- create_sbm(edges, nodes)
my_sbm %>% show() %>% head()

```


Now that we have our model setup there are two parameters that control this merging. 

First we have `GREEDY`. When this is set to `TRUE`, the at every merge step the model will check every possible merge for each node and select the best one. 

Second is `N_CHECKS_PER_GROUP`. When `GREEDY` is set to `FALSE`, this parameter controls how many possible merges the model explores for each node before choosing the best one. Higher means a longer run time. Note that these potential merges are done by a random sample of neighbor groups so even if `N_CHECKS_PER_GROUP` is set to the total number of nodes in the network we are not guarenteed to explore all options like `GREEDY` does. 

We will perform a greedy merging here. 

```{r}
my_sbm$GREEDY <- TRUE
```

Now that we have our model and parameters set we can run the algorithm using the `collapse_groups()` method. This method takes two arguments, the first is the level to perform the agglomerative merge on (in this case the node-level or `0`), and the number of MCMC steps to perform in between each merge to equilibriate the model.

```{r}
# Run initialization algorithm on base level nodes
merge_results <- my_sbm$collapse_groups(0L, 1L)
```

This method returns a list containing the entropy and state of the model at each merge step. We can use these to look at how the merges progressed. 


First we will grab the number of groups at each merge step. To do this we pull out the state element and find how many nodes in it have a level of 1 (aka group level).


```{r}
extract_n_groups <- function(merge_results){
  merge_results %>%
  purrr::map('state') %>%
  purrr::map_int(~{
    .x %>%
      dplyr::filter(level == 1) %>%
      dplyr::pull(id) %>%
      length()
  })
}

extract_n_groups(merge_results) %>% head()
```


Next we can look at how that number of groups corresponds to the entropy of the model. 

```{r}
extract_entropy <- function(merge_results){
  merge_results %>% purrr::map_dbl('entropy')
}
extract_entropy(merge_results) %>% head()
```


We should almost always see the entropy decrease as we remove groups but if there is signal in the model and the algorithm found it, we should see a relativel "lack of decrease" around the true structure value. We know the true number of groups in this data is 6 so we can reference that here...

```{r}
results_to_df <- function(merge_results){
  tibble(
    n_groups = extract_n_groups(merge_results),
    entropy = extract_entropy(merge_results)
  )
}



plot_merge_results <- function(merge_results_df, group_cutoff = 30, trim_axes = TRUE){
  
  results_below_cutoff <- merge_results_df %>% 
    filter(n_groups < group_cutoff)
  
  max_entropy <- max(results_below_cutoff$entropy)
  min_entropy <- min(results_below_cutoff$entropy)
  
  p <- ggplot(merge_results_df, aes(x = n_groups, y = entropy)) +
    geom_line(alpha = 0.3) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = K_a + K_b, color = 'orangered')
    
  if(trim_axes){
    p <- p +
      coord_cartesian(
        xlim = c(0, group_cutoff),
        ylim = c(min_entropy, max_entropy)
      ) 
  }
  
  p +
    scale_x_reverse() +
    theme_minimal() +
    labs(x = "number of groups")
}
```


```{r}
merge_results %>% results_to_df() %>% plot_merge_results()
```


We can perform the same setup but this time not using greedy but the sampling version
```{r}
new_sbm <- create_sbm(edges, nodes)
new_sbm$GREEDY <- FALSE
new_sbm$N_CHECKS_PER_GROUP <- 10
non_greedy_merge_results <- new_sbm$collapse_groups(0L, 15L)

non_greedy_merge_results %>% results_to_df() %>% plot_merge_results()
```

We can try scanning over different number of checks per group to see how it changes results

```{r}
merges_over_changes <- purrr::map_dfr(
  seq(5, 50, by = 5),
  ~{
    new_sbm <- create_sbm(edges, nodes)
    new_sbm$GREEDY <- FALSE
    new_sbm$N_CHECKS_PER_GROUP <- .x
    merge_results <- new_sbm$collapse_groups(0L, 15L)
    results_to_df(merge_results) %>% mutate(checks = .x)
  })

plot_merge_results(merges_over_changes) + 
  facet_wrap(~checks) +
  ggtitle("Tuning number of merge checks per node")
```



```{r}
merges_over_sweeps_and_greedy <- expand.grid(
  mcmc_sweeps = seq(1, 101, by = 25),
  greedy = c(TRUE, FALSE)
) %>% 
  {
    purrr::map2_dfr(
      .$mcmc_sweeps,
      .$greedy,
      ~{
        new_sbm <- create_sbm(edges, nodes)
        new_sbm$GREEDY <- .y
        merge_results <- new_sbm$collapse_groups(0L, as.integer(.x))
        results_to_df(merge_results) %>% 
          mutate(mcmc_sweeps = .x,
                 greedy = paste("greedy = ", .y))
      }
    )
  }


plot_merge_results(merges_over_sweeps_and_greedy) + 
  facet_grid(mcmc_sweeps~greedy) +
  ggtitle("Tuning and greedy status and number of mcmc sweeps per stage")
```


```{r}
purrr::map_dfr(
  seq(from = 0.9, to = 2, length.out = 5),
  ~{
    new_sbm <- create_sbm(edges, nodes)
    new_sbm$GREEDY <- FALSE
    new_sbm$N_CHECKS_PER_GROUP <- 10
    new_sbm$SIGMA <- .x
    merge_results <- new_sbm$collapse_groups(0L, 15L)
    results_to_df(merge_results) %>% mutate(sigma = .x)
  }) %>% 
  plot_merge_results(trim_axes = FALSE) + 
  facet_wrap(~sigma) +
  ggtitle("Tuning sigma value for merging")
```

