---
title: "bipartite_networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bipartite_newtorks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

is_html <- knitr::is_html_output()

html_message <- if(is_html) "" else "_If you were building this in an HTML output this would be an interactive network plot_"


```

```{r setup}
library(sbmr)
library(dplyr)
library(tidyr)
library(ggplot2)
```



Start with the package included pollinators network.

```{r load-data}
clements_pollinators %>% head()

bind_rows(
  count(clements_pollinators, pollinator) %>% 
    transmute(id = pollinator, degree = n, type = "pollinator"),
  count(clements_pollinators, flower) %>% 
    transmute(id = flower, degree = n, type = "flower")
) %>% 
  ggplot(aes(x = degree)) +
  geom_histogram(binwidth = 1, color = 'white') +
  facet_grid(rows = "type", scales = "free_y") +
  labs(title = "Degree distributions for flower and pollinator nodes",
       subtitle = "Note different y-axis scales")
```

## Load into SBM

```{r}
pollinator_net <- new_sbm_network(edges = clements_pollinators,
                                  bipartite_edges = TRUE,
                                  edges_from_column = pollinator,
                                  edges_to_column = flower,
                                  random_seed = 42)

pollinator_net
```

## Visualize it

`r html_message`

```{r}
if(is_html){
  visualize_network(pollinator_net, 
                    node_color_col = 'type', 
                    node_shape_col = 'none')
}
```



## Run agglomerative merging to find initial MCMC State

```{r}
pollinator_net <- pollinator_net %>% 
  collapse_blocks(desired_n_blocks = 4,
                  num_mcmc_sweeps = 5,
                  sigma = 1.1)
```


```{r}
visualize_collapse_results(pollinator_net, heuristic = "delta_ratio")
```
So we have a very clear peak. Let's select it as our starting position. 

```{r}
pollinator_net <- pollinator_net %>% 
  choose_best_collapse_state(heuristic = "delta_ratio", verbose = TRUE)
```


Let's visualize this result on the network layout



`r html_message`


```{r}
if(is_html){
  pollinator_net %>% 
    visualize_network(node_shape_col = 'type', 
                      node_color_col = 'block')
}
```

## MCMC Sweeping

We can now run 200 sweeps from the initial position chosen by the collapse...

```{r}
num_sweeps <- 200

pollinator_net <- pollinator_net %>%
  mcmc_sweep(num_sweeps = num_sweeps, 
             eps = 0.1, 
             track_pairs = TRUE)

pollinator_net %>% 
  visualize_mcmc_trace() +
  labs(title = glue::glue('Result of {num_sweeps} MCMC sweeps'),
       subtitle = "Entropy Delta of sweep and number of nodes moved for sweep")
```


## Visualizing propensity matrix

```{r}
pollinator_net %>% 
  get_sweep_pair_counts() %>% 
  ggplot(aes(x = node_a, y = node_b, fill = proportion_connected)) +
  geom_raster() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(x = "node a", 
       y = "node b",
       fill = "Proportion of sweeps\nin same block",
       title = "Node block consensus matrix")
```

## ECDF of propensity distribution

```{r}
visualize_propensity_dist(pollinator_net)
```


`r html_message`

```{r}
if(is_html){
  pollinator_net %>% 
    visualize_propensity_network(proportion_threshold = 0.01, 
                                 isolate_type = "flower")
}
```




